Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2016-11-21 writeup/local_pca_paper.tex
Moved the reordering of the text to this branch while keeping correct figures.

As reviewed above,
it is common to describe variation along the genome of simple statistics such as $F_{ST}$,
and also to use methods such as PCA to visualize large-scale patterns in genome-wide relatedness.
In this paper we 
apply PCA locally along the genome,
thus describing in an unbiased, descriptive way
how patterns of mean relatedness vary systematically along the genome, 
in a way particularly suited to large samples from geographically distributed populations.
Geographic population structure sets the stage by establishing similar patterns of relatedness across much of the genome;
we then aim to describe how this structure is affected by selection and other factors.
Our aim is not to identify outlier loci,
but rather to describe larger-scale variation shared by many parts of the genome.
By doing this with three taxa with diverse population histories,
we can compare relative contributions of different sorts of variation across taxa;
since the method is an visualization method, 
we allow ourselves to be suprised by unexpected signals in the data.

@@

As depicted in \Figure \ref{fig:diagram}, the general steps to the method are:
(1) divide the genome into windows,
(2) summarize the patterns of relatedness in each window,
(3) measure dissimilarity in relatedness between each pair of windows,
(4) visualize the resulting dissimilarity matrix using multidimensional scaling (MDS),
and (5) combine similar windows to more accurately visualize local effects of population structure using PCA.
Details of how we carried out each step are given in the Methods,
and code to run the analyses is provided as an R package at \url{https://github.com/petrelharp/local_pca}.

We applied the method to genomic datasets with good geographic sampling:
380 African \textit{Drosophila melanogaster} from the Drosophila Genome Nexus \citep{lack2015drosophila},
a worldwide dataset of humans,
3,965 humans from several locations worldwide from the POPRES dataset \citep{nelson2008population},
and 263 \textit{Medicago truncatula} from 24 countries around the Mediterranean basin 
a range-wide dataset of the partially selfing weedy annual plant 
from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},

In all three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
This implies that variation is due to meaningful heterogeneity in a biological process,
since noise due to randomness in choice of local genealogical trees
is not expected to show long distance correlations. 
Below, we discuss the results and likely underlying causes.

@@

%%%%%%%%%%%%%%%%%
\section{Methods}

We first recoded sampled genotypes as numeric matrices in the usual manner,
by recording the number of nonreference alleles seen at each locus for each sample.
We then divided the genome into contiguous segments 
(``windows'')
% recoded matrix into contiguous matrices that have the same columns but fewer rows than the original matrix, 
and applied principal component analysis (PCA) as described in \citet{mcvean2009genealogical}
separately to the submatrices that corresponded to each window.
The choice of window length entails a tradeoff between signal and noise,
since shorter windows allow better resolution along the genome but provide less precise estimates of relatedness.
A method for choosing a window length to balance these considerations is given in Appendix \ref{apx:window_length}.
% Specifically, we did PCA as follows:
Precisely,
denote by $Z$ the $L\times N$ recoded genotype matrix for a given window ($L$ is the number of SNPs and $N$ is the sample size), 
and by $\overline{Z_{s}}$ the mean of non-missing entries for allele $s$, 
so that $\overline{Z_{s}}=\frac{1}{n_s}\sum_j Z_{sj}$, 
where the sum is over the $n_s$ nonmissing genotypes.
We first compute the mean-centered matrix $X$, as $X_{si}=Z_{si}-\overline{Z_{s}}$,
and preserving missingness.
(This mean-centering makes the result not depend on the choice of reference allele,
exactly if there is no missing data, and approximately otherwise.)
Next, we find the covariance matrix of $X$, denoted $C$,
as $C_{ij} = \frac{1}{m_{ij}-1} \sum_s X_{si} X_{sj} - \frac{1}{m_{ij}(m_{ij}-1)} (\sum_s X_{si})(\sum_s X_{sj})$,
where all sums are over the $m_{ij}$ sites where both sample $i$ and sample $j$ have nonmissing genotypes.
% We compute the covariance matrix using the R function cov() , with use=``pairwise",
% which computes the covariance between each pair of individuals using all complete pairs of SNPs on those individuals.
The principal components are the eigenvectors of $C$, 
normalized to have Euclidean length equal to one,
and ordered by magnitude of the eigenvalues.

We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
However, 
we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
since rotations or reflections of these imply identical patterns of relatedness.
Instead, we compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs,
which is invariant under reflections and rotations and yet contains all other information about the PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
Furthermore, to remove the effect of artifacts such as mutation rate variation,
we also rescale each approximate covariance matrix 
so that the underlying data matrix has trace norm equal to one.

The three publicly available datasets we used are summarized in Table \ref{tab:data_stats}.
We converted each to a numeric matrix
(with one row per polymorphic variant and one column per sample)
by replacing each genotype with the number of nonreference alleles
(or NA for missing data).
% A normalization step (see below) ensures the result does not depend on the choice of reference allele.

\section{Data access}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human}
    \end{center}
    \caption{
        PCA plots for all 22 huan autosomes from the POPRES data.
        \label{sfig:pca_human_allchr}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{drosophila_recomb_mds_correlation}
    \end{center}
    \caption{
        Recombination rate, and the effects of population structure for \textit{Drosophila melanogaster}:
        this shows the first MDS coordinate and recombination rate (in cM/Mbp), as in Supplementary \Figure \ref{fig:drosophila_recomb_rate},
        against each other.
        Since the windows underlying estimates of Supplementary \Figure \ref{fig:drosophila_recomb_rate} do not coincide,
        to obtain correlations we divided the genome into 100Kbp bins, 
        and for each variable (recombination rate and MDS coordinate 1)
        averaged the values of each overlapping bin with weight proportional to the proportion of overlap.
        The correlation coefficient and $p$-values for each linear regression are as follows:
        2L: correlation $=0.52$, $r^2=0.27$;
        2R: correlation $=0.43$, $r^2=0.18$;
        3L: correlation $=0.47$, $r^2=0.21$;
        3R: correlation $=0.46$, $r^2=0.21$;
        X:  correlation $=0.50$, $r^2=0.24$.
        \label{sfig:drosophila_gene_density_correlations}
    }
\end{figure}

>>>>

As reviewed above,
it is common to describe variation along the genome of simple statistics such as $F_{ST}$,
and also to use methods such as PCA to visualize large-scale patterns in genome-wide relatedness.
In this paper we 
apply PCA locally along the genome,
thus describing in an unbiased, descriptive way
how patterns of mean relatedness vary systematically along the genome, 
in a way particularly suited to large samples from geographically distributed populations.
Geographic population structure sets the stage by establishing similar patterns of relatedness across much of the genome;
we then aim to describe how this structure is affected by selection and other factors.
Our aim is not to identify outlier loci,
but rather to describe larger-scale variation shared by many parts of the genome.
By doing this with three taxa with diverse population histories,
we can compare relative contributions of different sorts of variation across taxa;
since the method is an visualization method, 
we allow ourselves to be surprised by unexpected signals in the data.

%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

As depicted in \Figure \ref{fig:diagram}, the general steps to the method are:
(1) divide the genome into windows,
(2) summarize the patterns of relatedness in each window,
(3) measure dissimilarity in relatedness between each pair of windows,
(4) visualize the resulting dissimilarity matrix using multidimensional scaling (MDS),
and (5) combine similar windows to more accurately visualize local effects of population structure using PCA.
% Details of how we carried out each step are given in the Methods,
% and code to run the analyses is provided as an R package at \url{https://github.com/petrelharp/local_pca}.

To begin,
we first recoded sampled genotypes as numeric matrices in the usual manner,
by recording the number of nonreference alleles seen at each locus for each sample.
We then divided the genome into contiguous segments 
(``windows'')
% recoded matrix into contiguous matrices that have the same columns but fewer rows than the original matrix, 
and applied principal component analysis (PCA) as described in \citet{mcvean2009genealogical}
separately to the submatrices that corresponded to each window.
The choice of window length entails a tradeoff between signal and noise,
since shorter windows allow better resolution along the genome but provide less precise estimates of relatedness.
A method for choosing a window length to balance these considerations is given in Appendix \ref{apx:window_length}.
% Specifically, we did PCA as follows:
Precisely,
denote by $Z$ the $L\times N$ recoded genotype matrix for a given window ($L$ is the number of SNPs and $N$ is the sample size), 
and by $\overline{Z_{s}}$ the mean of non-missing entries for allele $s$, 
so that $\overline{Z_{s}}=\frac{1}{n_s}\sum_j Z_{sj}$, 
where the sum is over the $n_s$ nonmissing genotypes.
We first compute the mean-centered matrix $X$, as $X_{si}=Z_{si}-\overline{Z_{s}}$,
and preserving missingness.
(This mean-centering makes the result not depend on the choice of reference allele,
exactly if there is no missing data, and approximately otherwise.)
Next, we find the covariance matrix of $X$, denoted $C$,
as $C_{ij} = \frac{1}{m_{ij}-1} \sum_s X_{si} X_{sj} - \frac{1}{m_{ij}(m_{ij}-1)} (\sum_s X_{si})(\sum_s X_{sj})$,
where all sums are over the $m_{ij}$ sites where both sample $i$ and sample $j$ have nonmissing genotypes.
% We compute the covariance matrix using the R function cov() , with use=``pairwise",
% which computes the covariance between each pair of individuals using all complete pairs of SNPs on those individuals.
The principal components are the eigenvectors of $C$, 
normalized to have Euclidean length equal to one,
and ordered by magnitude of the eigenvalues.

We think of the local effects of population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
However, 
we do not compare patterns of relatedness of different genomic regions by directly comparing the PCs,
since rotations or reflections of these imply identical patterns of relatedness.
Instead, we compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs,
which is invariant under reflections and rotations and yet contains all other information about the PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
Furthermore, to remove the effect of artifacts such as mutation rate variation,
we also rescale each approximate covariance matrix to be of similar size
(precisely, so that the underlying data matrix has trace norm equal to one).

We applied the method to genomic datasets with good geographic sampling:
380 African \textit{Drosophila melanogaster} from the Drosophila Genome Nexus \citep{lack2015drosophila},
a worldwide dataset of humans,
3,965 humans from several locations worldwide from the POPRES dataset \citep{nelson2008population},
and 263 \textit{Medicago truncatula} from 24 countries around the Mediterranean basin 
a range-wide dataset of the partially selfing weedy annual plant 
from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
as summarized in Table \ref{tab:data_stats}.

\subsection{Data access}

@@

In all three datasets:
a worldwide sample of humans, African \textit{Drosophila melanogaster},
and a rangewide sample of \textit{Medicago truncatula},
PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
This implies that variation is due to meaningful heterogeneity in a biological process,
since noise due to randomness in choice of local genealogical trees
is not expected to show long distance correlations. 
Below, we discuss the results and likely underlying causes.

@@

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{FigS_pca_plot_allchr_human}
    \end{center}
    \caption{
        PCA plots for all 22 human autosomes from the POPRES data.
        \label{sfig:pca_human_allchr}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{drosophila_recomb_mds_correlation}
    \end{center}
    \caption{
        Recombination rate, and the effects of population structure for \textit{Drosophila melanogaster}:
        this shows the first MDS coordinate and recombination rate (in cM/Mbp), as in \Figure \ref{fig:drosophila_recomb_rate},
        against each other.
        Since the windows underlying estimates of \Figure \ref{fig:drosophila_recomb_rate} do not coincide,
        to obtain correlations we divided the genome into 100Kbp bins, 
        and for each variable (recombination rate and MDS coordinate 1)
        averaged the values of each overlapping bin with weight proportional to the proportion of overlap.
        The correlation coefficient and $p$-values for each linear regression are as follows:
        2L: correlation $=0.52$, $r^2=0.27$;
        2R: correlation $=0.43$, $r^2=0.18$;
        3L: correlation $=0.47$, $r^2=0.21$;
        3R: correlation $=0.46$, $r^2=0.21$;
        X:  correlation $=0.50$, $r^2=0.24$.
        \label{sfig:drosophila_gene_density_correlations}
    }
\end{figure}
